\begin{description}

\item[Debuggee]
	The piece of software for which we want to analyze cache usage. The term debuggee is used because it’s run with valgrind which is usually used for debugging.
\item[Matrix]
	A matrix is, in our case, an area of the memory which the debuggee has marked to contain data that is to be analyzed by our software. Access An access is an event generated for the debuggee. It marks loading or storing data to the main memory with a specific address.
\item[Relative Access]
	We are matching memory accesses to matrices and compute coordinates from that. For each access and it's previous access, a difference is coputed, and that is called a relative access. \newline
	\begin{tikzpicture} %Tikzpictures are going to be fun. Our coordinate systems diverge...
		\draw [color=gray] (-0.5,-0.5) grid (1.5,1.5);
		\node (frst) at (1,0) {}; %nth access
		\node (scnd) at (0,1) {}; %n+1-th access
		\draw [->] (frst) -- (scnd) node [midway,right,draw=none] {$(-1,-1)$};
	\end{tikzpicture}
\item[Pattern]
	A pattern is an ordered set of relative accesses. These accesses have been observed subsequently multiple (to be exact: at least two) times. It is not guaranteed that the first access of a pattern always occures first. Patterns can't tell you anything about when or in which compositions they occur
\item[Sequence]
	When scanning for patterns in the access stream, a series of consecutive accesses is often accounted to one pattern. The length of this series and what happens next is bundled into a sequence.
\end{description}
